---
title: "PA1_template"
output: html_document
---
## 1. Code for reading in the dataset and/or processing the data

```{r setup}
# load libraries
library(magrittr)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(devtools)
library(lubridate)
# download file from web
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile = "activity.zip", mode="wb")
# unzip data and read 
unzip("activity.zip")
stepdata <- read.csv("activity.csv", header = TRUE)
# data transformations
stepdata$date <- ymd(stepdata$date)
    stepdata$weekend <- as.factor(ifelse(weekdays(stepdata$date)=="Saturday" | weekdays(stepdata$date)=="Sunday","weekend","weekday"))
    stepdata$dayofweek <- as.factor(weekdays(stepdata$date))
# visually inspect data
head(stepdata)
```

## 2. Histogram of the total number of steps taken each day


```{r stepsperday}
databydate <- stepdata %>% select(date, steps) %>% group_by(date) %>% summarize(tsteps= sum(steps)) %>%na.omit()
hist(databydate$tsteps, xlab = "Total Daily Steps",main="Histogram of Total Steps by Day", breaks = 20)
```

## 3. Calculate and report the mean and median of the total number of steps taken per day


```{r meanmedian}
mean(databydate$tsteps)
median(databydate$tsteps)
```

## 4. Time series plot of the average number of steps taken

```{r plotavgsteps}
databyinterval <- stepdata%>% select(interval, steps) %>% na.omit() %>% group_by(interval) %>% summarize(tsteps= mean(steps)) 
ggplot(databyinterval, aes(x=interval, y=tsteps))+ geom_line()
```


## 5. The 5-minute interval that, on average, contains the maximum number of steps

```{r fivemininterval}
databyinterval[which(databyinterval$tsteps== max(databyinterval$tsteps)),]
```


## 6. Code to describe and show a strategy for imputing missing data

```{r missvals}
sum(is.na(stepdata))
# subset dataset where there are no NAs
    activity_no_NA <- stepdata[which(!is.na(stepdata$steps)),]
  
  # calculate the mean steps for each interval
    interval_only <- activity_no_NA %>% group_by(interval) %>% summarise(average=mean(steps))

    # convert the average to integer
    interval_only$average <- as.integer(interval_only$average)
    
    #subset dataset where steps have NAs
    activity_na <- stepdata[which(is.na(stepdata$steps)),]
    
    # fill NAs with average steps based on interval
    activity_na$steps <- ifelse(activity_na$interval==interval_only$interval,interval_only$average)
    
    # row bind the datasets that do not have NAs and the dataset where NAs are replaced with
    # mean values
    activity_impute <- rbind(activity_no_NA,activity_na)
```
    
## 7. Histogram of the total number of steps taken each day after missing values are imputed

```{r newnist}
  stepsByDay_impute <- activity_impute %>% group_by(date) %>% summarise(stepsperday = sum(steps))
    qplot(stepsperday,data=stepsByDay_impute,na.rm=TRUE,binwidth=500,xlab='Total Steps per Day', ylab='Frequency',main = 'Histogram of the total number of steps taken each day')
      totalstepsperday_impute <- activity_impute %>% group_by(date) %>% summarise(stepsperday = sum(steps))
  mean_n_median <- totalstepsperday_impute %>% summarise(average=mean(stepsperday),median=median(stepsperday))
  mean_n_median
```

## 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r panelplot}
meansteps <- activity_impute %>% group_by(interval,weekend) %>%   summarise(average = mean(steps))
  qplot(interval,average,data=meansteps,geom="line",facets=weekend~.,xlab="5-minute interval",ylab="average number of steps",main="Average steps pattern between Weekday and Weekend")
```